---
title: 'PSYC 211: Final Project'
author: "Holly Huey, Srihita Rudraraju, Janna Wennberg"
date: "12/14/2020"
output: pdf_document
---

```{r setup, include=FALSE, echo=FALSE}
# Setup, load required packages
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(viridis)

# Create colorspace from viridis
colorspace = viridis(6)

# Set seed
set.seed(211)
```

## Define functions
```{r}
# replace: a function that converts musical data to numbers
# takes in a vector of letters as inputs
# 'O' denotes the C that is an octave up from C
replace = function(draws){
  draws_num = gsub('C', 1, draws)
  draws_num = gsub('D', 2, draws_num)
  draws_num = gsub('E', 3, draws_num)
  draws_num = gsub('F', 4, draws_num)
  draws_num = gsub('G', 5, draws_num)
  draws_num = gsub('A', 6, draws_num)
  draws_num = gsub('B', 7, draws_num)
  draws_num = gsub('O', 8, draws_num)
  return(draws_num)
}

# get_sequence: a function that generates a sequence of musical notes from a transition matrix
# takes in the transition matrix (data), song length (n), and the probabilities for choosing the starting note
get_sequence = function(data, n, starting_prob){
  out = numeric(n)
  out[1] = sample(notes, 1, replace=TRUE, prob=starting_prob) # Change to NULL to have it be truly uniform sampling
  for (i in 2:n){
    out[i] = sample(notes, 1, replace=TRUE, prob=data[out[i-1],])
  }
  return(out)
}

```

## Load and clean data
```{r}
# Nursery rhymes are presented below in alphabetical order
abc = as.numeric(replace(c("C", "C", "G", "G", "A", "A", "G", "F", "F", "E", "E", "D", "D", "C",
        "G", "G", "F", "E", "E", "D", "G", "G", "F", "E", "E", "D",
        "C", "C", "G", "G", "A", "A", "G", "F", "F", "E", "E", "D", "D", "C")))

away_in_a_manger = as.numeric(replace(c("C", "F", "F", "G", "A", "F", "F", "A", "C", "C", "C",
                     "D", "G", "A", "B", "B", "C", "A", "A", "F", "A", "G", "D", "F",
                     "A", "C", "F", "F", "G", "A", "F", "F", "A", "C", "C",
                     "D", "B", "G", "A", "B", "B", "C", "A", "A", "F", "A", "G", "D", "A", "F")))

black_sheep = as.numeric(replace(c('C','C','G','G','A','A','A','A','G',
                'F','F','E','E','D','D','C',
          'G','G','G','F','F','E','E','E','D',
          'G','G','G','F','F','F','F','E','E','E','D',
          'C','C','G','G','A','A','A','A','G',
          'F','F','E','E','D','D','C')))

bingo = as.numeric(replace(c('G','C','C','C','G','A','A',
          'G','G','C','C','D','D','E','C',
          'E','E','F','F','F','D','D',
          'E','E','E','C','C','D','D','D',
          'C','B','G','A','B','C','C')))

fails_do_do = as.numeric(replace(c("B", "A", "G", "G", "A", "G", "A", "B", "G", "B", "A", "G", "G", "A", "G", "A", "B", "G")))

flintstones = as.numeric(replace(c("G", "C", "O", "A", "G", "C", "G", "F", "E", "E", "F", "G", "C", "D", "E",
                "G", "C", "O", "A", "G", "C", "G", "F", "E", "E", "F", "G", "C", "D", "E",
                "B", "E", "O", "B", "B", "A", "A", "B", "A", "A", "D", "B", "A", "A", "G", "G", "A", "G",
                "G", "C", "O", "A", "G", "C", "G", "F", "E", "E", "F", "G", "C", "D", "E", "F", "G", "C", "D", "E", "F", "G", "O", "O")))

frere_jaques = as.numeric(replace(c('C','D','E','C',
                 'C','D','E','C',
                 'E','F','G',
                 'E','F','G',
                 'G','A','G','F','E','C',
                 'G','A','G','F','E','C',
                 'C','G','C','C','G','C')))

hey_diddle_diddle = as.numeric(replace(c("E", "E", "E", "E", "F", "G", "D", "D", "D", "D", "C", "D", "E", "E",
                      "E", "E", "F", "G", "D", "E", "F", "F", "F", "F", "G", "A", "G", "E",
                      "C", "D", "E", "G", "G", "G", "G", "A", "B", "O")))

hickory_dickory = as.numeric(replace(c("E", "F", "G", "F", "E", "D", "E", "E", "E", "G", "F", "D", "E", "E",
                  "E", "E", "G", "G", "F", "F", "A", "G", "A", "G", "F", "E", "D", "C")))

hot_cross_buns = as.numeric(replace(c('E','D','C','E','D','C',
                   'C','C','C','C',
                   'D','D','D','D',
                   'E','D','C')))

humpty_dumpty = as.numeric(replace(c('C','C','E','E','D','E',
                  'D','C','E','E','G','G', 
                  'F','G','F','E','A','A',
                  'A','G','G','G','F','F',
                  'F','E','F','F','F','E',
                  'E','E','D','E','D','C')))

i_love_you_barney = as.numeric(replace(c("G", "E", "G", "G", "E", "G", "A", "G", "F", "E", "D", "E", "F",
                      "E", "F", "G", "C", "C", "C", "C", "C", "D", "E", "F", "G",
                      "G", "D", "D", "F", "E", "D", "C", "G", "E", "G", "G", "E", "G",
                      "A", "G", "F", "E", "D", "E", "F", "E", "F", "G", "C", "C", "C",
                      "C", "C", "D", "E", "F", "G", "G", "D", "D", "F", "E", "D", "C")))

im_a_little_teapot = as.numeric(replace(c("C", "D", "E", "F", "G", "O", "A", "O", "G", "F", "F", "F", "E", "E",
                       "D", "D", "D", "C", "C", "D", "E", "F", "G", "O", "A", "O", "G", "O",
                       "A", "G", "G", "F", "E", "D", "C")))

itsy_bitsy_spider = as.numeric(replace(c("G", "C", "C", "C", "D", "E", "E",
                      "E", "D", "C", "D", "E", "C", "E", "E", "F", "G",
                      "G", "F", "E", "F", "G", "E", "C", "C", "D", "E",
                      "E", "D", "C", "D", "E", "C", "G", "G", "C", "C",
                      "C", "D", "E", "E", "E", "D", "C", "D", "E", "C")))

jingle_bells = as.numeric(replace(c("E", "E", "E", "E", "E", "E", "E", "G", "C", "D", "E", "F", "F",
                 "F", "F", "F", "E", "E", "E", "E", "D", "D", "E", "D", "G")))

london_bridge = as.numeric(replace(c('G','A','G','F','E','F','G',
                  'D','E','F','E','F','G',
                  'G','A','G','F','E','F','G',
                  'D','G','E','C')))

londons_burning = as.numeric(replace(c("C", "C", "F", "F", "C", "C", "F", "F", "G", "G", "A", "A",
                    "G", "G", "A", "A", "C", "C", "C", "C", "C", "B", "A", "A", "C", "B", "A", "A")))

los_pollitos_dicen = as.numeric(replace(c("C", "D", "E", "F", "G", "G", "A", "A", "A", "A", "G", "G",
                       "F", "F", "F", "G", "E", "E", "D", "D", "D", "E", "C", "C")))

mary_lamb = as.numeric(replace(c('E','D','C','D','E','E','E',
              'D','D','D','E','G','G',
              'E','D','C','D','E','E','E',
              'E','D','D','E','D','C')))

ode_to_joy = as.numeric(replace(c("E", "E", "F", "G", "G", "F", "E", "D", "C", "C", "D", "E", "E", "D", "D",
               "E", "E", "F", "G", "G", "F", "E", "D", "C", "C", "D", "E", "D", "C", "C",
               "D", "D", "E", "C", "D", "E", "F", "E", "C", "D", "E", "F", "E", "D", "C", "D", "G",
               "E", "E", "F", "G", "G", "F", "E", "D", "C", "C", "D", "E", "E", "D", "D")))

old_macdonald = as.numeric(replace(c('C','C','C','G','A','A','G',
                  'E','E','D','D','C',
                  'G','C','C','C','G','A','A','G',
                  'E','E','D','D','C')))

rain_rain_go_away = as.numeric(replace(c('G','E','G','G','E',
                      'G','G','E','A','G','G','E',
                      'F','F','D','D','F','F','D',
                      'G','F','E','D','E','C','C')))

ring_around_the_rosie = as.numeric(replace(c("G", "G", "E", "A", "G", "E", "F", "G", "G", "G", "E",
                          "A", "G", "E", "F", "D", "F", "D", "E", "G", "E", "C")))

row_your_boat = as.numeric(replace(c("C", "C", "C", "D", "E", "E", "D", "E", "F", "G", "O", "O",
                  "O", "G", "G", "G", "E", "E", "E", "C", "C", "C", "G", "F", "E", "D", "C")))

this_old_man = as.numeric(replace(c('G','E','G','G','E','G',
                 'A','G','F','E','D','E','F',
                 'E','F','G','C','C','C','C',
                 'C','D','E','F','G',
                 'G','D','D','F','E','D','C')))

three_blind_mice = as.numeric(replace(c("E", "D", "C", "E", "D", "C", "G", "F", "F", "E", "G", "F", "F",
                     "E", "G", "O", "O", "B", "A", "B", "O", "G", "G", "G", "O",
                     "O", "O", "B", "A", "B", "O", "G", "G", "G", "G", "O", "O",
                     "B", "A", "B", "O", "G", "G", "G", "G", "E", "D", "C")))
                 
twinkle_twinkle = as.numeric(replace(c('C','C','G','G','A','A','G',
                    'F','F','E','E','D','D','C', 
                    'G','G','F','F','E','E','D', 
                    'G','G','F','F','E','E','D', 
                    'C','C','G','G','A','A','G',
                    'F','F','E','E','D','D','C')))

wheels_on_the_bus = as.numeric(replace(c("C", "F", "F", "F", "F", "A", "O", "A", "F", "G", "E", "C",
                      "A", "G", "F", "C", "F", "F", "F", "F", "A", "O", "A", "F", "G", "C", "C", "F")))

yankee_doodle = as.numeric(replace(c("C", "C", "D", "E", "C", "E", "D", "G", "C", "C", "D", "E", "C", "B",
                  "G", "C", "C", "D", "E", "F", "E", "D", "C", "B", "G", "A", "B", "C",
                  "C", "A", "B", "A", "G", "A", "B", "C", "G", "A", "G", "F", "E", "G",
                  "A", "B", "A", "G", "A", "B", "C", "A", "G", "C", "B", "D", "C", "C")))



# Create a list with the vectors loaded above
# CHECK: make sure all songs are present in this list
song_list = list(abc,away_in_a_manger,black_sheep,bingo,fails_do_do,flintstones,
                 frere_jaques,hey_diddle_diddle,hickory_dickory,hot_cross_buns,
                 humpty_dumpty,i_love_you_barney,im_a_little_teapot,itsy_bitsy_spider,
                 jingle_bells,london_bridge,londons_burning,los_pollitos_dicen,mary_lamb,
                 ode_to_joy,old_macdonald,rain_rain_go_away,ring_around_the_rosie,row_your_boat,
                 this_old_man,three_blind_mice,twinkle_twinkle,wheels_on_the_bus,yankee_doodle)

# Combine all vectors
# This combined vector will be used in two places: first to check the total number notes,
# and second to get the relative frequency of each note, which was used to determine
# the starting note in each song.
songs_comb = c(abc,away_in_a_manger,black_sheep,bingo,fails_do_do,flintstones,
                 frere_jaques,hey_diddle_diddle,hickory_dickory,hot_cross_buns,
                 humpty_dumpty,i_love_you_barney,im_a_little_teapot,itsy_bitsy_spider,
                 jingle_bells,london_bridge,londons_burning,los_pollitos_dicen,mary_lamb,
                 ode_to_joy,old_macdonald,rain_rain_go_away,ring_around_the_rosie,row_your_boat,
                 this_old_man,three_blind_mice,twinkle_twinkle,wheels_on_the_bus,yankee_doodle)

# Remove individual songs for housekeeping purposes
rm(abc,away_in_a_manger,black_sheep,bingo,fails_do_do,flintstones,
   frere_jaques,hey_diddle_diddle,hickory_dickory,hot_cross_buns,
   humpty_dumpty,i_love_you_barney,im_a_little_teapot,itsy_bitsy_spider,
   jingle_bells,london_bridge,londons_burning,los_pollitos_dicen,mary_lamb,
   ode_to_joy,old_macdonald,rain_rain_go_away,ring_around_the_rosie,row_your_boat,
   this_old_man,three_blind_mice,twinkle_twinkle,wheels_on_the_bus,yankee_doodle)

```

# Modeling
## Create transition matrix using nursery rhymes above
```{r}
# Create transition matrix
notes = c(1:8)
tally = matrix(1,length(notes),length(notes))

for (i in 1:length(song_list)){
  for (j in 1:length(song_list[[i]])){
    if (j!=1){
      prevNote = song_list[[i]][j-1]
      curNote = song_list[[i]][j]
      tally[prevNote,curNote] = tally[prevNote,curNote]+1
    }
  }
}

# Normalize transition matrix
transition_mat = data.frame(tally)
for (i in 1:nrow(tally)){
  transition_mat[i,] = transition_mat[i,]/sum(transition_mat[i,])
}
```

## Generate songs using the transition matrix generated above
```{r}
# Calculate mean length of each nursery rhyme
song_lengths = lengths(song_list)
median_song_length = median(song_lengths)

# Calculate relative frequency overall
# These probabilities will be used to determine the starting note
starting_prob_unnorm = tabulate(songs_comb)
starting_prob_vals = starting_prob_unnorm / sum(starting_prob_unnorm)

# Generate 100 songs
n_songs = 100
generated_songs = matrix(0,nrow=median_song_length,ncol=n_songs)
generated_songs = data.frame(generated_songs)
for (i in 1:n_songs){
  generated_songs[,i] = get_sequence(transition_mat,median_song_length,starting_prob_vals)
}

# Now make transition matrix for generated songs
tally_gen = matrix(1,length(notes),length(notes))
for (i in 1:ncol(generated_songs)){
  for (j in 1:nrow(generated_songs)){
    if (j!=1){
      prevNote = generated_songs[i,j-1]
      curNote = generated_songs[i,j]
      tally_gen[prevNote,curNote] = tally_gen[prevNote,curNote]+1
    }
  }
}

# Normalize transition matrix for generated songs
transition_mat_gen = data.frame(tally_gen)
for (i in 1:nrow(tally_gen)){
  transition_mat_gen[i,] = transition_mat_gen[i,]/sum(transition_mat_gen[i,])
}
```

# Inference
## Get a null set of songs
```{r}
# Generate null set of songs by shuffling rows of the matrix
null_songs = matrix(0,nrow=median_song_length,ncol=n_songs)
null_songs = data.frame(null_songs)
for (i in 1:n_songs){
  null_songs[,i] = sample(notes,median_song_length,replace=TRUE)
}

# Now make transition matrix for null songs
null_tally_gen = matrix(1,length(notes),length(notes))
for (i in 1:ncol(null_songs)){
  for (j in 1:nrow(null_songs)){
    if (j!=1){
      prevNote = null_songs[i,j-1]
      curNote = null_songs[i,j]
      null_tally_gen[prevNote,curNote] = null_tally_gen[prevNote,curNote]+1
    }
  }
}

# Normalize transition matrix for null songs
null_transition_mat_gen = data.frame(null_tally_gen)
for (i in 1:nrow(tally_gen)){
  null_transition_mat_gen[i,] = null_transition_mat_gen[i,]/sum(null_transition_mat_gen[i,])
}
```

## Pitch Proximity Principle
### Get pitch proximities for generated songs and null songs
```{r}
# Calculate distance jumps in generated songs
distances = matrix(0,nrow=(median_song_length-1),ncol=n_songs)
distances = data.frame(distances)
for (i in 1:ncol(distances)){
  for (j in 1:nrow(distances)){
    distances[j,i] = generated_songs[j+1,i]-generated_songs[j,i]
  }
}

# Now do the same for null songs
null_distances = matrix(0,nrow=(median_song_length-1),ncol=n_songs)
null_distances = data.frame(null_distances)
for (i in 1:ncol(null_distances)){
  for (j in 1:nrow(null_distances)){
    null_distances[j,i] = null_songs[j+1,i]-null_songs[j,i]
  }
}

# Get a list of combined distances for both null and generated songs
distances_unlisted = unname(unlist(distances))
null_distances_unlisted = unname(unlist(null_distances))
distances_comb = data.frame(cbind(distances_unlisted,null_distances_unlisted))
colnames(distances_comb) = c("distance","distance_null")
```

### Do a hypothesis test
Results:
$F = 0.350$, $df = 3099, 3099$, $p < 2.2 \times 10^{-16}$, $conf = [0,0.371]$
Ratio of variances = 0.350
```{r}
# Test for pitch proximity principle
# var.test performs an F test to compare the variances of two samples from normal populations
var.test(distances_comb$distance,distances_comb$distance_null, ratio=1, alternative="less")

```

## Key Profile
### Code data as consonant or disonant
```{r}
# CONSONANCE
# Get a list of what the consonant bigrams are
cons = list(c(1,3), c(3,5), c(1,5), c(1,8),
            c(3,1), c(5,3), c(5,1), c(8,1))

# Use tally to determine the number of consonant transitions for generated songs
cons_indiv = c()
null_cons_indiv = c()
for (i in 1:length(cons)){
  cons_indiv[i] = tally_gen[cons[[i]][1], cons[[i]][2]] 
  null_cons_indiv[i] = null_tally_gen[cons[[i]][1], cons[[i]][2]]
}
n_cons = sum(cons_indiv)
total_bigrams = sum(tally_gen)
prop_cons = n_cons/total_bigrams

null_n_cons = sum(null_cons_indiv)
null_prop_cons = null_n_cons/total_bigrams

# DEVIL'S TRIAD (IV-VII)
# Use tally to determine the number of Devil's Triads for generated songs
n_dts = tally_gen[4,7]
null_n_dts = null_tally_gen[4,7]
```

### Do hypothesis tests
Results for consonance:
$\chi^2 = 19.51$, $df = 1$, $p = 5\times10^{-6}$
Prop 1 = 0.211, Prop 2 = 0.137

Results for Devil's Triad:
$\chi^2 = 11.25$, $df = 1$, $p = 0.0004$
Prop 1 = 0.0028, Prop 2 = 0.0189
```{r}
# Do a t-test to determine if a difference in proportion of consonant bigrams
prop.test(x=c(n_cons,null_n_cons),
          n=c(total_bigrams,total_bigrams),
          alternative=c("greater"))

# Do a t-test to determine if a difference in proportion of Devil's Triads
prop.test(x=c(n_dts,null_n_dts),
          n=c(total_bigrams,total_bigrams),
          alternative=c("less"))
```

# Plots
## Stacked bar chart
```{r}
# Stacked bar chart for generated songs
transition_mat_gen_long = expand.grid(c("C","D","E","F","G","A","B","High-C"),
                                      c("C","D","E","F","G","A","B","High-C"))
transition_mat_gen_long <- transition_mat_gen_long[ , c("Var2","Var1")]
colnames(transition_mat_gen_long) = c("startNote","nextNote")
for (i in 1:length(notes)){
  for (j in 1:length(notes)){
    transition_mat_gen_long$prob[length(notes)*(i-1)+j] = transition_mat_gen[i,j]
  }
}

transition_mat_gen_long %>%
  ggplot(aes(fill=nextNote,y=prob,x=startNote)) + 
  geom_bar(position="fill",stat="identity") +
  scale_fill_viridis(name="Next Note", discrete=TRUE)+
  xlab("Starting Note") + 
  ylab("Proportion")+
  theme_minimal()
ggsave("stacked_bar_chart.png")
```

## Distance histogram
```{r}
distances_comb %>%
  ggplot(aes(x=distance))+
  geom_histogram(binwidth=1,color="black",fill=colorspace[3])+
  theme_minimal()+
  theme(
    panel.grid.minor.y = element_blank()
  )+
  xlab("Distance")+
  ylab("Count")
ggsave("dist_hist.png")

# CHECK: the null histogram looks weirdly lopsided? won't change the major results but want to make sure...
# I think it's just a sampling thing...
distances_comb %>%
  ggplot(aes(x=distance_null))+
  geom_histogram(binwidth=1,color="black",fill=colorspace[3])+
  theme_minimal()+
  theme(
    panel.grid.minor.y = element_blank()
  )+
  xlab("Distance")+
  ylab("Count")
```

# Save out datasets
```{r}
# Save out original transition matrix
save(tally,file="saved_files/tally.Rda")
save(transition_mat,file="saved_files/transition_mat.Rda")
write.csv(transition_mat,file="saved_files/transition_mat.csv")

# Save generated songs and write to CSV
save(generated_songs,file="saved_files/generated_songs.Rda")
write.csv(generated_songs,file="saved_files/generated_songs.csv")

# Save transition matrix for generated songs
save(tally_gen,file="saved_files/tally_gen.Rda")
save(transition_mat_gen,file="saved_files/transition_mat_gen.Rda")
write.csv(transition_mat_gen,file="saved_files/transition_mat_gen.csv")

# Save null songs
save(null_songs,file="saved_files/null_songs.Rda")
write.csv(null_songs,file="saved_files/null_songs.csv")

# Save transition matrix for null songs
save(null_tally_gen,file="saved_files/null_tally_gen.Rda")
save(null_transition_mat_gen,file="saved_files/null_transition_mat_gen.Rda")
write.csv(null_transition_mat_gen,file="saved_files/null_transition_mat_gen.csv")

# Save distances
save(distances,file="saved_files/distances.Rda")
write.csv(distances,file="saved_files/distances.csv")
save(null_distances,file="saved_files/null_distances.Rda")
write.csv(null_distances,file="saved_files/null_distances.csv")
save(distances_comb,file="saved_files/distances_comb.Rda")
write.csv(distances_comb,file="saved_files/distances_comb.csv")
```




